EBASE manuscript

### 2023-03-22 manu-draft.qmd
more rgn edits, added nse methods

The additional modifications to @eq-base for EBASE were removing the $p$ and $\Theta$ parameters from the production (P) and respiration (R) components, respectively. The BASE method provides options to estimate both $\Theta$ and $p$ from the data or to set as fixed values defined in the model file ($\Theta = 1.07177$, $p=1$). For EBASE, the temperature dependency of respiration is removed by setting $\Theta^{(T-\bar{T})}$ = 1 and it is assumed that respiration is constant as in other methods [@Odum56; @Murrel18]. The $p$ parameter, which is dimensionless, is also set to 1 as in @Holtgrieve10, such that a linear relationship of production to $PAR$ was used. This removes the exponential dependence of production with increasing $PAR$. Removing these components also improved model stability by creating more reasonable estimates for the other parameters.

@@

The metabolism estimates and their parameters returned by the Bayesian routine implemented in JAGS are affected by the prior distributions assigned to each. As for the BASE model, relatively uninformed prior distributions following a normal Gaussian distribution are used by default, although the priors can be changed based on previous knowledge of parameters specific to an ecosystem or as informed by other metabolic modeling approaches. Reasonable uninformed prior distributions for EBASE were chosen based on normal distributions with mean values based on approximate estimates from the literature [@Wanninkhof14;@Grace15] and standard deviations that were sufficiently large to allow the Bayesian routine to search an unconstrained parameter space.  The default priors for $a$, $r$, and $b$ are:

@@

For ease of application, an R package was developed to implement the EBASE methods described above [@Beck22]. A detailed web page (<https://fawda123.github.io/EBASE/>) describes the required data inputs for estimating the metabolic parameters using the core R functions that pass inputs to the JAGS software. A JAGS model file is included with the package that implements @eq-ebase, with options to supply a custom model file as needed. Users can also specify alternative prior distributions from those in Equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior] for each parameter. Functions are available for viewing model results, including a plot of the modeled dissolved oxygen with the observed, a time series plot of the primary metabolic estimates (P, R, and D, as areal rates with units $\text{mmol}/\text{m}^3/\text{d}$), and a time series plot of the credible intervals for the $a$, $r$, and $b$ parameters. An additional option also allows the user to specify the model optimization period, where the time period can vary from a minimum of one day to the maximum of the entire length of the time series. This differs from the BASE approach where the optimization period is set at one day. Sensitivity of the EBASE method to different optimization periods is described in @sec-cmpmod. All results provided herein were generated using the EBASE R package.

@@

### EBASE comparison with modelled observations {#sec-cmpmod}

A preliminary assessment of the ability of EBASE to produce reasonable parameter estimates for $a$, $r$, and $b$ (@eq-ebase) was conducted using a synthetic time series with actual data as inputs. These assessments evaluated if 1) EBASE estimates similar values for the known parameters in the synthetic time series, and 2) how the estimated values change with different model configurations. The synthetic time series used inputs from an approximate one year of continuous data from Apalachicola Bay, Florida. The input data were 2012 observations at the Cat Point water quality monitoring station (29.7021$^{\circ}$ N, -84.8802$^{\circ}$ W) and included 15 minute observations of water temperature (C) and salinity (psu). Continuous meteorological observations from the nearby East Bay station (29.7909$^{\circ}$ N, -84.8834$^{\circ}$ W, \~10km north of Cat Point) were combined with the water quality data at the same time step and included air temperature (C), wind speed (m/s), and PAR (mmol/m$^2$ as total per 15 minute observation). Missing observations in the water quality and meteorological data were filled using autoregressive modelling of the actual parameters versus time to create a complete dataset from January 1st to December 31st. An oxygen time series was then generated using the inputs from Apalachicola Bay ([@fig-synapa]a), such that the synthetic dissolved oxygen was created using @eq-ebase as a forward calculating model and user-defined values for $a$, $r$, and $b$.  All metabolic rates for production, respiration, and gas exchange ([@fig-synapa]b and c) were also estimated with the synthetic time series. The only parameter that could not be created as a continuous time series was $b$, which was set to 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$ [@Wanninkhof14].  Accordingly, the synthetic dissolved oxygen time series and the observed water quality and meteorological data were used with EBASE to determine if reasonable estimates for the known parameters and metabolic rates were recovered.

Several configurations of the EBASE model can influence how the parameters and metabolic rates are estimated. Two types of sensitivity analyses were conducted to evaluate changes in the results returned by EBASE, where in all cases, the results were compared to the known values in the synthetic time series. First, results were compared to the known values after changing characteristics of their prior distributions. Second, results were compared using a model period of optimization for one day and seven days to assess how results varied based on the amount of data used for the Bayesian estimation. Details of these analyses are as follows.

@@

The optimization period used for a time series determines how much data are used to estimate the three EBASE parameters ($a$, $r$, and $b$). Unlike BASE, the EBASE software allows flexibility in the period of time used for optimization, with a minimum being one day and a maximum being the length of the time series. For a chosen optimization period, single estimates for $a$, $r$ and $b$ are returned by the model. A short period of optimization (e.g., one day) may return unstable parameter estimates if forcing factors (e.g., wind, temperature) that affect estuarine biogeochemical rates occur at time scales longer the chosen period of optimization for EBASE. As such, potentially longer optimization periods may return more stable and robust estimates given that more data are used for parameter estimation. An optimization period of one day was compared to results using an optimization period of seven days for the synthetic one-year time series. Each optimization period was evaluated for every unique combination of prior distributions, totaling 128 (64 times the two optimization periods) unique analyses for the assessment of changing priors of all parameters.

Changing the prior distributions affected the ability of EBASE to return the known parameters from the synthetic time series, with results varying somewhat by optimization period. @fig-priorcomp shows a summary of the comparisons of the synthetic data with EBASE for each unique combination of prior distributions and two optimization periods of one and seven days. The results are shown using Root Mean Squared Errors (RMSE) values comparing the synthetic estimates with those from EBASE. Darker colors represent higher RMSE values and can be used to assess how changing the priors affects the ability of EBASE to estimate the true parameter.

@@

### EBASE comparison with real observations

@@

-   Describe how EBASE deals with gaps

>>>>

The additional modifications to @eq-base for EBASE were removing the $p$ and $\Theta$ parameters from the production (P) and respiration (R) components, respectively. The BASE method provides options to estimate both $\Theta$ and $p$ from the data or to set as fixed values defined in the model file ($\Theta = 1.07177$, $p=1$). For EBASE, the temperature dependency of respiration is removed by setting $\Theta^{(T-\bar{T})}$ = 1 and it is assumed that respiration is constant as in other methods [@Odum56; @Murrell18]. The $p$ parameter, which is dimensionless, is also set to 1 as in @Holtgrieve10, such that a linear relationship of production to $PAR$ was used. This removes the exponential dependence of production with increasing $PAR$. Removing these components also improved model stability by creating more reasonable estimates for the other parameters.

@@

The metabolism estimates and their parameters returned by the Bayesian routine implemented in JAGS are affected by the prior distributions assigned to each. As for the BASE model, relatively uninformed prior distributions following a normal Gaussian distribution are used by default, although the priors can be changed based on previous knowledge of parameters specific to an ecosystem or as informed by other metabolic modeling approaches. Reasonable uninformed prior distributions for EBASE were chosen with mean values using approximate estimates from the literature [@Wanninkhof14;@Grace15] and standard deviations that were sufficiently large to allow the Bayesian routine to search an unconstrained parameter space.  The default priors for $a$, $r$, and $b$ are:

@@

For ease of application, an R package was developed to implement the EBASE methods described above [@Beck22]. A detailed web page (<https://fawda123.github.io/EBASE/>) describes the required data inputs for estimating the metabolic parameters using the core R functions that pass inputs to the JAGS software. A JAGS model file is included with the package that implements @eq-ebase, with options to supply a custom model file as needed. Users can also specify alternative prior distributions from those in Equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior] for each parameter. Functions are available for viewing model results, including a plot of the modeled dissolved oxygen with the observed, a time series plot of the primary metabolic estimates (P, R, and D, as areal rates with units $\text{mmol}/\text{m}^3/\text{d}$), and a time series plot of the credible intervals for the $a$, $r$, and $b$ parameters. An additional option allows the user to specify the model optimization period, where the time period can vary from a minimum of one day to the maximum of the entire length of the time series. This differs from the BASE approach where the optimization period is set at one day. Sensitivity of the EBASE method to different optimization periods is described in @sec-cmpmod. All results provided herein were generated using the EBASE R package.

@@

### EBASE comparison with known results {#sec-cmpmod}

A preliminary assessment of the ability of EBASE to produce reasonable parameter estimates for $a$, $r$, and $b$ (@eq-ebase) was conducted using a synthetic time series with actual data as inputs. These assessments evaluated if 1) EBASE estimates similar values for the known parameters in the synthetic time series, and 2) how the estimated values change with different model configurations. The synthetic time series used inputs from one year of continuous data from Apalachicola Bay, Florida. The input data were 2012 observations at the Cat Point water quality monitoring station (29.7021$^{\circ}$ N, --84.8802$^{\circ}$ W) and included 15 minute observations of water temperature and salinity. Continuous meteorological observations from the nearby East Bay station (29.7909$^{\circ}$ N, --84.8834$^{\circ}$ W, \~10km north of Cat Point) were combined with the water quality data at the same time step and included air temperature, wind speed, and PAR. Missing observations in the water quality and meteorological data were filled using autoregressive modelling of the actual parameters versus time to create a complete dataset from January 1st to December 31st. A synthetic oxygen time series was then generated using the inputs from Apalachicola Bay ([@fig-synapa]a) and @eq-ebase as a forward calculating model with user-defined values for $a$, $r$, and $b$.  All metabolic rates for production, respiration, and gas exchange ([@fig-synapa]b and c) were also estimated with the synthetic time series. The only parameter that could not be created as a continuous time series was $b$, which was set to 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$ [@Wanninkhof14].  Accordingly, the synthetic dissolved oxygen time series and the observed water quality and meteorological data were used with EBASE to determine if reasonable estimates for the known parameters and metabolic rates were recovered.

Several configurations of the EBASE model can influence how the parameters and metabolic rates are estimated. Two types of sensitivity analyses were conducted to evaluate changes in the results returned by EBASE, where in all cases, the results were compared to the known values in the synthetic time series. First, results were compared to the known values after changing characteristics of their prior distributions. Second, results were compared using a model period of optimization for one, seven, and 30 days to assess how results varied based on the amount of data used for the Bayesian estimation. Details of these analyses are as follows.

@@

The optimization period used for a time series determines how much data are used to estimate the three EBASE parameters ($a$, $r$, and $b$). Unlike BASE, the EBASE software allows flexibility in the period of time used for optimization, with a minimum being one day and a maximum being the length of the time series. For a chosen optimization period, single estimates for $a$, $r$ and $b$ are returned by the model. A short period of optimization (e.g., one day) may return unstable parameter estimates if forcing factors (e.g., wind, temperature) that affect estuarine biogeochemical rates occur at time scales longer the chosen period of optimization for EBASE. As such, potentially longer optimization periods may return more stable and robust estimates given that more data are used for parameter estimation, although at the risk of overgeneralizing metabolic rates if variation occurs at time scales less than the optimization period. Results from optimization periods of one, seven, and thirty days were compared for the synthetic one-year time series. Each optimization period was also evaluated for every unique combination of prior distributions, totaling 192 (64 times the three optimization periods) unique analyses for the assessment of changing priors of all parameters.

Changing the prior distributions affected the ability of EBASE to return the known parameters from the synthetic time series, with results varying by optimization period. @fig-priorcomp shows a summary of the comparisons of the synthetic data with EBASE for each unique combination of prior distributions and the three optimization periods of one, seven, and thirty days. The results are shown using Nash-Sutcliffe Efficiency (NSE) values [@Nash70;@Moriasi07]:

$$
NSE = 1 - \frac{\sum\limits_{i=1}^n \left(Y_i^{Syn} - Y_i^{EBASE}\right)^2}{\sum\limits_{i=1}^n \left(Y_i^{Syn} - \bar{Y}^{Syn}\right)^2}
$$ {#eq-nse}

where $Y_i$ is the estimate from EBASE or the synthetic times series for the optimization period $i$, evaluated for each metabolic parameter, each unique combination of priors, and each optimization period. The value $n$ is the number of optimization periods in the one-year time series, i.e., $n=365$ for one day, $n \approx 52$ for seven days, and $n \approx 12$ for 30 days. All metabolic estimates were averaged within the number of days for each optimization period prior to calculating $NSE$ values, if appropriate. The $NSE$ value is conceptually similar to the coefficient of determination, but varies from $-\infty$ to 1.  Values in the positive range from 0 to 1 are desirable with 1 being a perfect fit, whereas values in the negative range indicate the mean of the synthetic data is a better predictor than EBASE.

<!-- edit -->

@@

<!-- edit -->
### EBASE comparison with real observations

@@

-   Describe how EBASE deals with gaps and flexibility of the model to accommodate diff optimization periods
-   b is sketchy, how to improve? Either fix or develop JAGS model further to include additional data, e.g. tidal height.
